### 2017-05-11 paper/paper.tex

\documentclass[9pt,letterpaper]{extarticle}
\usepackage{extsizes}
\usepackage[paper=letterpaper,margin=1in]{geometry}

% Table coloring: Must be imported before TikZ
\usepackage[table,x11names,dvipsnames]{xcolor}

\usepackage{xspace,amsmath,amsfonts,amssymb,hyperref,tikz,multirow}
\usepackage{graphicx}
% \graphicspath{{../data/R_plots/}}

\usepackage{parskip}

\newcommand{\todo}[1]{%
	\mbox{}% prevent marginpar from being on previous paragraph
	\marginpar{%
		\colorbox{red!80!black}{\textcolor{white}{to-do}}%
		\vspace*{-22pt}% hack!
	}%
	\textcolor{red}{#1}%
}

\usepackage{algorithmicx}
\usepackage{algorithm}
\usepackage[noend]{algpseudocode}

\usepackage{graphicx}
\usepackage{multicol}

\title{N-dimensional Tic Tac Toe, and Adventure in Modules}
\author{Alex Grasley \and Jeff Young \and Michael McGirr}
\date{}

\begin{document}
	\maketitle
	% \begin{abstract}
  %   Do we want/need an abstract?
	% \end{abstract}

  \section{Introduction}
  \section{Overview of Project}
  % A high-level overview of what your project is about. This should answer the
  % following questions: What is the domain? Who are the users? What kinds of
  % things can those users do with your project?

  \section{Program description}
  % A list and brief description of all of the signatures, modules, and functors
  % in your project. In the description of signatures and modules, you should
  % highlight especially important types and functions. The goal of this section
  % is to guide the reader through your source code, putting things into their
  % proper context. It isn’t necessary to describe every helper function, but this
  % section should provide a reasonably complete overview of the content of your
  % project.

	\section{Design Decisions}
  % (Most important!) Describe in detail 3-5 design decisions you made during your
  % project, with respect to abstraction and/or modularity. For example, what is a
  % piece of information that you chose to hide and why? How did you decide on a
  % particular way of separating your system into modules compared to another. How
  % did you reduce coupling between two modules? These design decisions should
  % emphasize how you applied ideas from the papers we’ve read (and associated
  % discussions) to the design of your project. It might be helpful here to
  % include snippets from old versions (or hypothetical alternative versions) of
  % your project to compare and contrast with the final version.
  \subsection{Creating an Abstract Game Engine}
  \subsection{Higher Ordered Signatures, and the ``Include'' incantation}
  \subsection{Separation of IO, or How I learned to not fight SML in search of Purity}
  \subsection{You can do it in 2-dimensions, but can you do it in n-dimensions!}
  \subsection{The Functor is love, the Functor is life}

\end{document}

### 2017-05-12 paper/paper.tex
Let's sprinkle some pretty words on this - initial paper commit

@@
 
+  Our project initially began as a pokemon simulator but early on we realized
+  that we could make better use of the SML module system by approaching the
+  problem of simulating player-based games in a more abstract way.
+  By doing this we could use the basic notion of what a game simulation required
+  and isolate a pattern to follow for any number of games that fit this
+  adversarial model.
+  A game would then be a specific implementation - in our case tic tac toe -
+  that used this pattern.
+
   \section{Program description}
@@
   % project.
+  Our project is separated between the code that describes the game simulation
+  and the code that used this to make a specific tic tac toe implementation.
+  The game simulation is located in the \texttt{game.sml} file. The code for 
+  tic tac toe implementation is spread out between \texttt{tictactoe.sml} and
+  \texttt{matrix.sml}.
 

### 2017-05-12 paper/paper.tex
Adding more to section 3

--- 2017-05-11 paper/paper.tex
@@
 
+  Our project initially began as a pokemon simulator - but early on we realized
+  that we could make better use of the SML module system by approaching the
+  problem of simulating player-based games in a more abstract general way.
+  By doing so we could define the basic notion of what a game simulation requires
+  and isolate a pattern to follow for any number of games that fit this
+  model.
+  A game would then be a specific implementation - in our case
+  \textit{Tic-tac-toe} - that used this pattern.
+
   \section{Program description}
@@
   % project.
+
+  Our project is separated between the code that describes the game simulation
+  and the code that uses this to make a specific \textit{Tic-tac-toe} implementation.
+  The game simulation is located in the \texttt{game.sml} file. Likewise, the
+  code for the \textit{Tic-tac-toe} implementation is located in
+  \texttt{tictactoe.sml} and uses modules from \texttt{matrix.sml}.
+
+  As we touched on before - the notion of a game is generalized in the
+  \texttt{game.sml} file. This defines the modules to run a game.
+  The general pattern for a game under our model consists of three purposely
+  isolated pieces.
+  These are \textit{State}, \textit{Actions}, and \textit{Agents}.
+  These are then wrapped together with a functor. The idea is that any game
+  consists of a state, a set of actions on that state, and an agents that take
+  actions for a state. A specific implementation of a game would use the
+  relationship between these three general pieces to define a runnable game.
+
+  Our implementation of \textit{Tic-tac-toe} uses this game model to operate and
+  is primarily defined within \texttt{tictactoe.sml}.
+  Here we have two signatures - one that extends state (the \texttt{STATE}
+  signature from \texttt{game.sml}) for \textit{Tic-tac-toe}
+  called \texttt{TTTSTATE} and another that does the same with action called
+  \texttt{TTTACTION} - we don't extend the \texttt{AGENT} signature specifically.
+
+  There are many module structures instantiated here for the various kinds of
+  state, actions and agents that we eventually want to use -
+  such as \texttt{TttState} and \texttt{TttAction} for instance.
+  These are bound to functors like \texttt{TttStateFn} which takes
+  a module implementing a square matrix module (\texttt{SQUAREMATRIX})
+  and returns a module that implements \texttt{TTTSTATE}.
+
+  The functor \texttt{TttActionFn} (used by modules structures like
+  \texttt{TttAction} and \texttt{Ttt3DAction}) takes a module that implements
+  \texttt{TTTSTATE} and gives us a module implementing \texttt{TTTACTION}.
+
+  The idea being that we created specific modules structures
+  (like \texttt{TttState}) for different
+  cases of actions and states that may occur in various kinds of
+  \textit{Tic-tac-toe} games. We then described functors
+  which linked these structures to the functions meant for that specific use.
+  When provided with the correct input module these functors will give us
+  modules that implement that portion of the game pattern - which we then use
+  with another structure and another functor for the next modular piece of the
+  game.
+  % ugh I should probably rewrite that last paragraph
 

### 2017-05-12 paper/paper.tex
More work on section 3 - I think I'm going to rework the last paragraph and then move on to later sections

@@
   As we touched on before - the notion of a game is generalized in the
-  \texttt{game.sml} file. This defines the modules to run a game.
+  \texttt{game.sml} file. This defines the signatures and functor to run a game.
   The general pattern for a game under our model consists of three purposely
   isolated pieces.
-  These are \textit{State}, \textit{Actions}, and \textit{Agents}.
+  These are \textit{State}, \textit{Actions}, and \textit{Agents} which are
+  each given their own signature in \texttt{game.sml}.
   These are then wrapped together with a functor. The idea is that any game
@@
   actions for a state. A specific implementation of a game would use the
-  relationship between these three general pieces to define a runnable game.
+  relationship between these three general pieces to define a runnable game with
+  its own modules.
 
@@
   called \texttt{TTTSTATE} and another that does the same with action called
-  \texttt{TTTACTION} - we don't extend the \texttt{AGENT} signature specifically.
+  \texttt{TTTACTION} - we don't extend the \texttt{AGENT} signature specifically
+  for \textit{Tic-tac-toe}.
 
@@
   \texttt{TTTSTATE} and gives us a module implementing \texttt{TTTACTION}.
+  The functors \texttt{TttRandomAgent} and \texttt{TttHumanAgentFn} take
+  modules which implement \texttt{TTTACTION} and return a module that implements
+  \texttt{AGENT}.
 
-  The idea being that we created specific modules structures
-  (like \texttt{TttState}) for different
-  cases of actions and states that may occur in various kinds of
-  \textit{Tic-tac-toe} games. We then described functors
-  which linked these structures to the functions meant for that specific use.
+  We can then instantiate a structure like \texttt{TttExecRandom} that will run
+  a game of \textit{Tic-tac-toe} by providing our \texttt{ExecFn} functor from
+  \texttt{game.sml} with a module that implements \texttt{AGENT}.
+  \texttt{ExecFn} then gives us a module that implements \texttt{EXEC}.
+
+  The idea being that we instantiated specific module structures
+  (like \texttt{TttState}) for different cases of actions and states that may
+  occur in various kinds of \textit{Tic-tac-toe} games.
+  We then described functors which linked these structures to the functions
+  meant for that specific use and returned a new .
   When provided with the correct input module these functors will give us

### 2017-05-12 paper/paper.tex
Section 3 is mostly done - I didn't talk about the repl, matrix or utils files very much. I'm going to start poking around with the Design Decision section now.

@@
   As we touched on before - the notion of a game is generalized in the
-  \texttt{game.sml} file. This defines the signatures and functor to run a game.
+  \texttt{game.sml} file. This defines the signatures which make up the pattern
+  of a game and a functor to run a game.
   The general pattern for a game under our model consists of three purposely
   isolated pieces.
-  These are \textit{State}, \textit{Actions}, and \textit{Agents} which are
+  These are \textit{State}, \textit{Actions}, and \textit{Agents} - which are
   each given their own signature in \texttt{game.sml}.
-  These are then wrapped together with a functor. The idea is that any game
-  consists of a state, a set of actions on that state, and an agents that take
-  actions for a state. A specific implementation of a game would use the
-  relationship between these three general pieces to define a runnable game with
-  its own modules.
+  These are then wrapped together with a functor to execute a game.
+  The idea is that any game consists of a state, a set of actions on that state,
+  and an agents that take actions for a state. A specific implementation of a
+  game would use the relationship between these three general pieces to define
+  a runnable game with its own modules.
 
@@
   is primarily defined within \texttt{tictactoe.sml}.
-  Here we have two signatures - one that extends state (the \texttt{STATE}
-  signature from \texttt{game.sml}) for \textit{Tic-tac-toe}
-  called \texttt{TTTSTATE} and another that does the same with action called
-  \texttt{TTTACTION} - we don't extend the \texttt{AGENT} signature specifically
-  for \textit{Tic-tac-toe}.
+  This roughly follows the order:
+  State $\rightarrow$ Action $\rightarrow$ Agent $\rightarrow$ Execution.
 
-  There are many module structures instantiated here for the various kinds of
-  state, actions and agents that we eventually want to use -
-  such as \texttt{TttState} and \texttt{TttAction} for instance.
-  These are bound to functors like \texttt{TttStateFn} which takes
-  a module implementing a square matrix module (\texttt{SQUAREMATRIX})
+  In \texttt{tictactoe.sml} we have two signatures - one that extends state
+  (the \texttt{STATE} signature from \texttt{game.sml}) for \textit{Tic-tac-toe}
+  called \texttt{TTTSTATE} and another that does the same with action
+  (the \texttt{ACTION} signature from \texttt{game.sml}) called
+  \texttt{TTTACTION}.
+  We don't extend the \texttt{AGENT} signature specifically for \textit{Tic-tac-toe}.
+
+  There are many module structures instantiated in \texttt{tictactoe.sml}
+  for the various kinds of state, actions and agents that we eventually want to
+  use - such as \texttt{TttState} and \texttt{TttAction} for instance - but each will
+  be used to implement the next kind of module.
+  These module structures are bound to functors like \texttt{TttStateFn} which
+  takes a module implementing a square matrix module (\texttt{SQUAREMATRIX})
   and returns a module that implements \texttt{TTTSTATE}.
@@
   The idea being that we instantiated specific module structures
-  (like \texttt{TttState}) for different cases of actions and states that may
-  occur in various kinds of \textit{Tic-tac-toe} games.
+  (like \texttt{TttState}) for different cases of states, actions and agents
+  that may occur in various kinds of \textit{Tic-tac-toe} games.
   We then described functors which linked these structures to the functions
-  meant for that specific use and returned a new .
-  When provided with the correct input module these functors will give us
-  modules that implement that portion of the game pattern - which we then use
-  with another structure and another functor for the next modular piece of the
-  game.
-  % ugh I should probably rewrite that last paragraph
+  meant for that specific use and returned a new module.
+  When provided with the correct input module, these functors will give us
+  modules that implement that portion of the overall game pattern.
 

### 2017-05-12 paper/paper.tex
Add latest edits to section 3

@@
   \section{Introduction}
+
   \section{Overview of Project}
@@
   takes a module implementing a square matrix module (\texttt{SQUAREMATRIX})
-  and returns a module that implements \texttt{TTTSTATE}.
+  and returns a module that implements \texttt{TTTSTATE}. The functors in
+  \texttt{tictactoe.sml} effectively take a module structure that implements the
+  previous module of the game pattern and produce the current (or next)
+  modular piece.
 
@@
 
-  We can then instantiate a structure like \texttt{TttExecRandom} that will run
-  a game of \textit{Tic-tac-toe} by providing our \texttt{ExecFn} functor from
+  We can then instantiate a structure (like \texttt{TttExecRandom} for example)
+  that will run a game of \textit{Tic-tac-toe}
+  To do so we provide our \texttt{ExecFn} functor from
   \texttt{game.sml} with a module that implements \texttt{AGENT}.

### 2017-05-12 paper/paper.tex
IO section done

@@
   \subsection{Separation of IO, or How I learned to not fight SML in search of Purity}
+  In the first implementation of our project we simulated Haskell's IO monad by
+  creating a ``show'' signature, and an ``IO'' functor, as shown below:
+
+  \begin{verbatim}
+    signature SHOW =
+    sig
+        type a
+        val show : a -> string
+    end
+
+    signature IO =
+    sig
+        structure S : SHOW
+
+        val printIO : S.a -> unit
+        val read : 'a -> string option
+        val say : string -> unit
+    end
+
+    functor Io (structure Sh : SHOW) : IO =
+    struct
+
+      structure S = Sh
+
+      (* append a new line to a str, this is expensive *)
+      fun appendNewLine str = implode $ (explode str) @ [#"\n"]
+
+      fun printIO x = print o appendNewLine o S.show $ x
+
+      (* function to get user input, it doesn't do anything with its argument *)
+      fun read _ = TextIO.inputLine TextIO.stdIn
+
+      fun say str = print o appendNewLine $ str
+
+    end
+  \end{verbatim}
+  We decided against this approach i.e. isolating IO into a separate module,
+  because it began to pollute our design's dependency graph with unwanted edges.
+  In haskell type class instances exist in an global overloaded name space, that
+  takes advantage of haskell's dispatch system. So each data type that is
+  implemented for the \texttt{Show} typeclass is able to be printed to StdOut
+  \textit{without} carrying around its respective show function. This is not the
+  case in SML and it was sorely missed in our implementation. If we were to
+  follow the Haskell style of IO (separating impure and pure code explicitly),
+  then an IO dependency would be requisite anytime we needed to perform any IO.
+  Thus, the IO node would become a dominating node in our dependency graph and
+  the import of it into every functor would quickly become unweidly and
+  boilerplate. Hence, we chose an SML style design where the structures that are
+  mapped upon carry any relevant functions with them to maintain a clean
+  decomposition in our design. This type of structure bloat, although
+  disgraceful to a haskell programming, seems to be encouraged by SML's module
+  system.
+
   \subsection{You can do it in 2-dimensions, but can you do it in n-dimensions!}

### 2017-05-12 paper/paper.tex
added n-dimensionality discussion, could use some word smithing

@@
 	% \end{abstract}
-
-  \section{Introduction}
 
@@
   \subsection{Creating an Abstract Game Engine}
+  % How we came up with the abstract design
   \subsection{Higher Ordered Signatures, and the ``Include'' incantation}
+  % Something about how previously we just had signature -> structure -> functors
+  % but with include it became possible to 
   \subsection{Separation of IO, or How I learned to not fight SML in search of Purity}
@@
   \subsection{You can do it in 2-dimensions, but can you do it in n-dimensions!}
+  Initially, we had conceived on implementing a pokemon battle simulator. As we
+  dove into the problem domain we realized that instead of falling into a
+  specific, flow-chart design with modules like: Pokemon, Combat, Moves etc. it
+  would be better design to create an abstract game engine and then make pokemon
+  \textit{an instance of} the abstract game engine. A similar pattern occurred
+  during our first implementation of the matrix library utilized by the tic tac
+  toe state viz. we realized we wrote a, too specific, flowchart-y instance of a
+  matrix library. Thus, in the spirit of modularity, we refactored the matrix
+  library to implement an abstract \texttt{CONTAINER} signature. Then based on
+  this signature, it became possible to create sub-signatures that specified the
+  types of the CONTAINER signature. A good example is shown below:
+
+  \begin{verbatim}
+    signature CONTAINER = sig
+
+      (* ADT of the container *)
+      type 'a container
+      (* type used to index the container *)
+      type index
+      (* type used to represent the size of the container *)
+      type size
+    ...
+    end
+
+    (* Matrices that are square and so only need size to be a single int *)
+    signature SQUAREMATRIX = sig
+
+      include CONTAINER where type size = int
+
+      val intToIndex : 'a container * int -> index
+
+    end
+
+  \end{verbatim}
+  In this example we show the relevant parts of the \texttt{CONTAINER} signature
+  and how we used that signature to create a \textit{more specific}
+  \texttt{SQUAREMATRIX} signature e.g. a matrix whose size datatype need only be
+  a single integer value. This allows for a very flexible and modular
+  implementation, in addition to square matrices we could create jagged matrices
+  by following the same pattern viz. defining a \texttt{JAGGEDMATRIX} and
+  setting size to ``int * int''.
+
+  Not only is the second layer of abstraction useful for changing matrix
+  properties on the fly, but it is also useful for changing the underlying
+  implementation and dimensionality of the matrix. With this design we were able
+  to create a functor that, when given a structure of type
+  \texttt{SQUAREMATRIX}, and a structure of type \texttt{VECT} will return a
+  matrix of whose underlying datatype is implemented by a structure, V of type
+  \texttt{VECT}, that dictates the underlying datatype of the matrix, and a
+  structure M, of type \texttt{SQUAREMATRIX} that specifies the datatype that
+  stores the V structure. For example, we create matrices that are created with
+  arrays, vectors and lists by passing each as a structure of type
+  \texttt{VECT}; any such data structure is viable as long as it supports the
+  \texttt{VECT} signature. By altering the structure M, of type
+  \texttt{SQUAREMATRIX} we create n-dimensional matrices. So by varying both we
+  gain the ability to create n-dimensional matrices, with abstracted underlying
+  data types. Thus, the double abstracted design of our matrix library is a
+  semantic definition of a matrix, which separates and abstracts the underlying
+  datatype implementation from the dimensionality of the matrix and the
+  underlying implementation details, thereby allowing for extreme reuse and
+  flexibility.
+
   \subsection{The Functor is love, the Functor is life}
+  % How functors enabled our clean decomposition and modular design
 

### 2017-05-12 paper/paper.tex
minor fixes

@@
   because it began to pollute our design's dependency graph with unwanted edges.
-  In haskell type class instances exist in an global overloaded name space, that
-  takes advantage of haskell's dispatch system. So each data type that is
+  In haskell, type class instances exist in an global overloaded name space,
+  that takes advantage of haskell's dispatch system. So each data type that is
   implemented for the \texttt{Show} typeclass is able to be printed to StdOut
   \textit{without} carrying around its respective show function. This is not the
-  case in SML and it was sorely missed in our implementation. If we were to
+  case in SML, and it was sorely missed in our implementation. If we were to
   follow the Haskell style of IO (separating impure and pure code explicitly),
   then an IO dependency would be requisite anytime we needed to perform any IO.
-  Thus, the IO node would become a dominating node in our dependency graph and
-  the import of it into every functor would quickly become unweidly and
-  boilerplate. Hence, we chose an SML style design where the structures that are
-  mapped upon carry any relevant functions with them to maintain a clean
-  decomposition in our design. This type of structure bloat, although
-  disgraceful to a haskell programming, seems to be encouraged by SML's module
-  system.
+  Thus, the IO node would become a dominating node in our dependency graph, and
+  the its import into every functor would quickly become boilerplate and
+  unweidly. Hence, we chose an SML style design, where the structures that are
+  mapped upon carry any relevant functions with them, which maintains a clean
+  decomposition in SML. This type of structure bloat, although disgraceful to a
+  haskell programming, seems to be encouraged by SML's module system.
 

### 2017-05-12 paper/paper.tex
editing for style

@@
 
-  As we touched on before - the notion of a game is generalized in the
+  As we touched on before, the notion of a game is generalized in the
   \texttt{game.sml} file. This defines the signatures which make up the pattern
@@
   The idea is that any game consists of a state, a set of actions on that state,
-  and an agents that take actions for a state. A specific implementation of a
+  and agents that select which action to perform on a given state. A specific implementation of a
   game would use the relationship between these three general pieces to define
@@
 
-  In \texttt{tictactoe.sml} we have two signatures - one that extends state
+  In \texttt{tictactoe.sml} we have two signatures: one that extends state
   (the \texttt{STATE} signature from \texttt{game.sml}) for \textit{Tic-tac-toe}
@@
   We don't extend the \texttt{AGENT} signature specifically for \textit{Tic-tac-toe}.
+  These extended signatures are perfectly compatible with the original signatures
+  due to the structural matching of SML signatures.
 
@@
   for the various kinds of state, actions and agents that we eventually want to
-  use - such as \texttt{TttState} and \texttt{TttAction} for instance - but each will
+  use---such as \texttt{TttState} and \texttt{TttAction} for instance---but each will
   be used to implement the next kind of module.
@@
 
-  The functor \texttt{TttActionFn} (used by modules structures like
+  The functor \texttt{TttActionFn} (used by module structures like
   \texttt{TttAction} and \texttt{Ttt3DAction}) takes a module that implements
   \texttt{TTTSTATE} and gives us a module implementing \texttt{TTTACTION}.
-  The functors \texttt{TttRandomAgent} and \texttt{TttHumanAgentFn} take
+  The functors \texttt{TttRandomAgentFn} and \texttt{TttHumanAgentFn} take
   modules which implement \texttt{TTTACTION} and return a module that implements
@@
   We can then instantiate a structure (like \texttt{TttExecRandom} for example)
-  that will run a game of \textit{Tic-tac-toe}
+  that will run a game of \textit{Tic-tac-toe}.
   To do so we provide our \texttt{ExecFn} functor from
@@
   \subsection{You can do it in 2-dimensions, but can you do it in n-dimensions!}
-  Initially, we had conceived on implementing a pokemon battle simulator. As we
+  Our initial intent was to implement a pokemon battle simulator. As we
   dove into the problem domain we realized that instead of falling into a
@@
   during our first implementation of the matrix library utilized by the tic tac
-  toe state viz. we realized we wrote a, too specific, flowchart-y instance of a
-  matrix library. Thus, in the spirit of modularity, we refactored the matrix
+  toe state. We wrote a single matrix module with more rigidly defined 
+  dimensions and underlying implementation details. While it was abstract, it was not very modular,
+  as it required writing a new module from the ground up if any of the 
+  implementation details were to be changed. Thus, in the spirit of modularity, we refactored the matrix
   library to implement an abstract \texttt{CONTAINER} signature. Then based on
@@
   a single integer value. This allows for a very flexible and modular
-  implementation, in addition to square matrices we could create jagged matrices
+  implementation, in addition to square matrices we could create jagged two-dimensional matrices
   by following the same pattern viz. defining a \texttt{JAGGEDMATRIX} and

### 2017-05-12 paper/paper.tex
Adding the abstract game design section

--- 2017-05-12 paper/paper.tex
@@
   % your project to compare and contrast with the final version.
+
   \subsection{Creating an Abstract Game Engine}
   % How we came up with the abstract design
+  The design for our game engine stemmed from trying to isolate the common
+  general components of what would be required to create a running game.
+  Rather than focus on creating one specific game implementation - we wanted to
+  look at the larger picture and view the problem in terms of the common pattern
+  that emerged between these kinds of games.
+
+  Each of the portions of the \textit{State}-\textit{Actions}-\textit{Agents}
+  design is independent so that - once created - it does not rely on other
+  modules for it's internal pieces.
+  Each of the constituent parts of the game engine is more akin to a node on a
+  graph and the interfaces are like edges.
+  The pieces should be thought of as islands of code that could then interact
+  with other modules. By using the game model to define a specific game - we
+  essentially create more specific versions of those islands following the
+  recipe spelled out in the signatures.
+
   \subsection{Higher Ordered Signatures, and the ``Include'' incantation}

### 2017-05-12 paper/paper.tex
Fix a typo and adding one more sentence

@@
   general components of what would be required to create a running game.
-  Rather than focus on creating one specific game implementation - we wanted to
+  Rather than focus on creating one specific game implementation where these
+  problems are addressed in an \textit{ad hoc} way - we wanted to
   look at the larger picture and view the problem in terms of the common pattern
@@
   design is independent so that - once created - it does not rely on other
-  modules for it's internal pieces.
+  modules for its internal pieces.
   Each of the constituent parts of the game engine is more akin to a node on a

### 2017-05-12 paper/paper.tex
structural typing section

--- 2017-05-12 paper/paper.tex
@@
   % your project to compare and contrast with the final version.
-
   \subsection{Creating an Abstract Game Engine}
   % How we came up with the abstract design
-  The design for our game engine stemmed from trying to isolate the common
-  general components of what would be required to create a running game.
-  Rather than focus on creating one specific game implementation - we wanted to
-  look at the larger picture and view the problem in terms of the common pattern
-  that emerged between these kinds of games.
-
-  Each of the portions of the \textit{State}-\textit{Actions}-\textit{Agents}
-  design is independent so that - once created - it does not rely on other
-  modules for it's internal pieces.
-  Each of the constituent parts of the game engine is more akin to a node on a
-  graph and the interfaces are like edges.
-  The pieces should be thought of as islands of code that could then interact
-  with other modules. By using the game model to define a specific game - we
-  essentially create more specific versions of those islands following the
-  recipe spelled out in the signatures.
-
-  \subsection{Higher Ordered Signatures, and the ``Include'' incantation}
+  \subsection{Exploiting Structural Typing}
   % Something about how previously we just had signature -> structure -> functors
   % but with include it became possible to
+  One major element of SML's module system that we were about to exploit in our project was
+  the structural matching of modules to signatures. Coupled with liberal use of the \texttt{include}
+  keyword, this allowed us to easily extend signature definitions in a way reminiscent of inheritance
+  and subclassing in OOP.
+
+  This design decision is most apparent in our extension of the abstract game engine to
+  our chosen implementation domain of \textit{Tic-tac-toe}. Our more abstract interfaces defined
+  in \texttt{game.sml} were often insufficient for describing the full interface
+  needed for our concrete implementation. For example, \textit{Tic-tac-toe} state modules
+  needed to export a \texttt{cell} datatype that represented the actual Xs and Os placed on the game
+  board. Including this datatype in the abstract \texttt{STATE} signature in \texttt{game.sml}
+  would be inappropriate, as any game outside of \textit{Tic-tac-toe} has no need for the
+  \texttt{cell} datatype. Instead, we created a new signature \texttt{TTTSTATE} that extended
+  the \texttt{STATE} signature via the \texttt{include} keyword with additional interface elements.
+  Additionally, this extension refined
+  some of the abstract types in \texttt{STATE} using the \texttt{where} keyword in order to give them
+  concrete type definitions. For example, the abstract type \texttt{state} is defined to always be
+  equivalent to the type \texttt{cell * (cell Matrix.container)} in a module implementing \texttt{TTTSTATE}.
+
+  We were able to exploit this refined state signature to narrow the domain of the \texttt{TttActionFn}
+  functor to only states appropriate for \textit{Tic-tac-toe}, rather than any abstract game state.
+  However, thanks to the structural matching properties of the SML module system, we can still
+  use a module implementing \texttt{TTTSTATE} wherever we would use a module implementing
+  \texttt{STATE}. In fact, because the matching is structural and not nominal, we get this property
+  for free, with no need to specify the relationship by name in the code. This allows us to still use
+  signatures and functors in our implementation that depend only on the more abstract 
+  definitions like the \texttt{AGENT} signature
+  or the \texttt{ExecFn} functor, which are only defined for the more abstract game engine.
+
   \subsection{Separation of IO, or How I learned to not fight SML in search of Purity}

### 2017-05-12 paper/paper.tex

@@
   % your project to compare and contrast with the final version.
+
   \subsection{Creating an Abstract Game Engine}
   % How we came up with the abstract design
+  The design for our game engine stemmed from trying to isolate the common
+  general components of what would be required to create a running game.
+  Rather than focus on creating one specific game implementation where these
+  problems are addressed in an \textit{ad hoc} way, we wanted to
+  look at the larger picture and view the problem in terms of the common pattern
+  that emerged between these kinds of games.
+
+  Each of the portions of the \textit{State}-\textit{Actions}-\textit{Agents}
+  design is independent so that - once created - it does not rely on other
+  modules for its internal pieces.
+  Each of the constituent parts of the game engine is more akin to a node on a
+  graph and the interfaces are like edges.
+  The pieces should be thought of as islands of code that could then interact
+  with other modules. By using the game model to define a specific game - we
+  essentially create more specific versions of those islands following the
+  recipe spelled out in the signatures.
+
 \subsection{Exploiting Structural Typing}

### 2017-05-12 paper/paper.tex
coupling section

@@
   because it began to pollute our design's dependency graph with unwanted edges.
-  In haskell, type class instances exist in an global overloaded name space,
-  that takes advantage of haskell's dispatch system. So each data type that is
+  In Haskell, type class instances exist in an global overloaded name space,
+  that takes advantage of Haskell's dispatch system. So each data type that is
   implemented for the \texttt{Show} typeclass is able to be printed to StdOut
@@
   the its import into every functor would quickly become boilerplate and
-  unweidly. Hence, we chose an SML style design, where the structures that are
+  unwieldy. Hence, we chose an SML style design, where the structures that are
   mapped upon carry any relevant functions with them, which maintains a clean
   decomposition in SML. This type of structure bloat, although disgraceful to a
-  haskell programming, seems to be encouraged by SML's module system.
+  Haskell programming, seems to be encouraged by SML's module system.
 
@@
   implementation details were to be changed. Thus, in the spirit of modularity, we refactored the matrix
-  library to implement an abstract \texttt{CONTAINER} signature. Then based on
-  this signature, it became possible to create sub-signatures that specified the
-  types of the CONTAINER signature. A good example is shown below:
+  library to implement a \texttt{CONTAINER} signature that specifies the interface for
+  a container abstract data type. As above, we were able to use structural matching and the
+  \texttt{include} keyword to extend our container ADT to more specific kinds of container. A good
+  example is shown below:
 
@@
   \texttt{SQUAREMATRIX} signature e.g. a matrix whose size datatype need only be
-  a single integer value. This allows for a very flexible and modular
+  a single integer value. Similarly, we can define a \texttt{VECT} signature that describes
+  containers that can be used as one-dimensional vectors (like list or arrays) by restricting 
+  both the \texttt{size} and \texttt{index} types to be \texttt{int}.
+  This allows for a very flexible and modular
   implementation, in addition to square matrices we could create jagged two-dimensional matrices
@@
 
-  Not only is the second layer of abstraction useful for changing matrix
-  properties on the fly, but it is also useful for changing the underlying
-  implementation and dimensionality of the matrix. With this design we were able
-  to create a functor that, when given a structure of type
-  \texttt{SQUAREMATRIX}, and a structure of type \texttt{VECT} will return a
-  matrix of whose underlying datatype is implemented by a structure, V of type
-  \texttt{VECT}, that dictates the underlying datatype of the matrix, and a
-  structure M, of type \texttt{SQUAREMATRIX} that specifies the datatype that
-  stores the V structure. For example, we create matrices that are created with
-  arrays, vectors and lists by passing each as a structure of type
-  \texttt{VECT}; any such data structure is viable as long as it supports the
-  \texttt{VECT} signature. By altering the structure M, of type
-  \texttt{SQUAREMATRIX} we create n-dimensional matrices. So by varying both we
-  gain the ability to create n-dimensional matrices, with abstracted underlying
-  data types. Thus, the double abstracted design of our matrix library is a
-  semantic definition of a matrix, which separates and abstracts the underlying
-  datatype implementation from the dimensionality of the matrix and the
-  underlying implementation details, thereby allowing for extreme reuse and
-  flexibility.
+  One consequence of this design is that you can create n-dimensional square matrices given only
+  modules implementing \texttt{VECT} and a functor. We define a functor \texttt{SquareMatrixFn}
+  that takes a module \texttt{M} implementing \texttt{SQUAREMATRIX} and a module \texttt{V}
+  implementing \texttt{VECT} and produces a new square matrix of dimension n+1 where n is the
+  dimensionality of M. We simply use \texttt{V} to store the polymorphic matrix elements and \texttt{M} to
+  store rows of \texttt{V}. The type of indices of our new matrix is just the product of the index types
+  of its components.
 
-  \subsection{The Functor is love, the Functor is life}
-  % How functors enabled our clean decomposition and modular design
+  Because all \texttt{VECT} modules also implement \texttt{SQUAREMATRIX}, we
+  can create two-dimensional matrices by passing any two \texttt{VECT} implementations to our functor.
+  From there, we can create n-dimensional matrices by simply passing the result back into the functor
+  as \texttt{M} any number of times. Therefore, given only a module implementing \texttt{VECT}, our
+  functor can automatically produce modules representing matrices of n-dimensions.
+  Because functors are not first-class, we cannot automate this process by passing in
+  the intended dimensionality as an integer argument, so instead we are stuck threading the results
+  of our functor back into itself. This could prove tedious for matrices with high dimensionality, but for our
+  purposes was acceptable.
+
+  This use of functors also allows for mixing of containers in matrices.
+  We implemented \texttt{VECT} for the basis array, list, and vector libraries. Our functor can take any
+  combination of these and produce a matrix module that works the same
+  as any other. As these implementation details are hidden by the container ADT, we can safely mix
+  any of them to form another valid container matrix.
+
+  All of this has the result that our \textit{Tic-tac-toe} implementation works for matrices of any
+  dimensionality and implemented using a variety of data structures.
+
+  \subsection{Coupling}
+
+  One area where our implementation encountered some issues was in the coupling of our system.
+  Despite our best efforts, our modules are not as loosely coupled as we would like. Each module
+  carries with it the modules that it depends on as part of its structure. This is because many of the
+  interfaces depend on type definitions contained in the modules that they depend on (for example,
+  \texttt{ACTION} modules depend on the definition of the \texttt{state} type defined in \texttt{STATE}).
+  This is not ideal in terms of
+  coupling, and forced us to use a lot of trivial functors that thread these dependencies through our
+  system.
+
+  Resolving this is a tricky problem. If we make the types of our abstract game interface all concrete,
+  we could gain looser coupling at the cost of how generic our abstract game engine can be. Concrete
+  type definitions in all signatures would alleviate this problem, but we would wind up unable to extend
+  our abstract interface to any other games aside from \textit{Tic-tac-toe}. Being locked into a single
+  implementation in this way would simply reduce our modules to simple namespaces with none of the 
+  abstraction or extensibility benefits associated with modular programming.
+
+  One possible solution to this problem could be to somehow serialize interactions between
+  different modules into a standard format (much like how web servers and clients often convert data
+  into JSON to pass back and forth). However, this still seems to include some notion of coupling, just
+  deferred now to serialization and deserialization processes. A module would still have to know how to
+  serialize and deserialize information and how to interpret the communications it receives from 
+  other modules. So while we would decouple at the type level, we would still retain coupling at the
+  informational level. It is unclear if there is a cleaner decomposition of our game domain that would
+  allow for looser coupling while still retaining similar abstraction and cohesion as our current system.
 

### 2017-05-13 paper/paper.tex
minor fixes

@@
 
-\title{N-dimensional Tic Tac Toe, and Adventure in Modules}
+\title{N-dimensional Tic Tac Toe, an Adventure in Modules}
 \author{Alex Grasley \and Jeff Young \and Michael McGirr}
@@
   Thus, the IO node would become a dominating node in our dependency graph, and
-  the its import into every functor would quickly become boilerplate and
+  then its import into every functor would quickly become boilerplate and
   unwieldy. Hence, we chose an SML style design, where the structures that are
@@
   decomposition in SML. This type of structure bloat, although disgraceful to a
-  Haskell programming, seems to be encouraged by SML's module system.
+  Haskell programmer, seems to be encouraged by SML's module system, and is
+  discussed in detail below.
 
